---
title: Massachusetts 2024 Election Overview
subtitle: "A detailed overview of primary and general races"
author: "Brent Benson"
date: "2024-08-04"
categories: [elections, politics, mapoli]
execute:
  echo: false
format:
  html:
    html-table-processing: none
---

```{r}
#| output: false

library(tidyverse)
library(readxl)
library(glue)
library(gt)

# Color of group-row
group_color <- "lightgrey"

# Leave out registrars and clerks
offices_to_cover <- c(
    "Governor's Council",
    "State Representative",
    "State Senate",
    "Representative in Congress",
    "Senator in Congress"
)      

candidates <- read_excel("2024 Candidates 6.4.24.xlsx") %>%
    # Column transformations
    mutate(is_incumbent = (`Incumbent Y/N` == "Y"),
           district = str_replace(District, " & ", " and "),
           candidate = str_to_title(`Candidate`),
                      street_address = str_split_i(Address, ", ", 1),
           city_town = str_split_i(Address, ", ", 2),
           display_name = str_glue("{candidate} ({Party}-{city_town})")) %>%
    # Renaming and selection of columns to keep
    select(office = Office,
           district,
           district_id = `District ID`,
           candidate,
           display_name,
           is_incumbent,
           party = Party,
           street_address,
           city_town,
           statement = `Statement of Public Office/Political Designation`)

office_list <- unique(candidates$office)

calculated_incumbents <- read_csv("ma_incumbents_2024_08_03.csv") %>%
    mutate(district = str_replace(district, " & ", " and "),
           incumbent_party = str_sub(party_incumbent, 1, 1),
           incumbent_display = str_glue("{name_incumbent} ({incumbent_party}-{city_town_incumbent})")) %>%
    select(office,
           district,
           incumbent = name_incumbent,
           incumbent_display,
           incumbent_party,
           incumbent_city_town = city_town_incumbent)

running_incumbents <- candidates %>%
    filter(is_incumbent) %>%
    select(office,
           district,
           incumbent=candidate,
           incumbent_display=display_name,
           incumbent_party=party,
           incumbent_city_town=city_town)

# Give the set of candidates for a particular district,
# return a dataframe with a column per party with the
# count of the candidates in that party.
party_counts <- function(candidates) {
    candidates %>%
        group_by(party) %>%
        tally() %>%
        pivot_wider(names_from=party, values_from=n)
}

district_summaries <- read_excel("notable_districts_2024.xlsx") %>%
    filter(!is.na(summary)) %>%
    select(district_id, summary)

districts <- candidates %>%
    # Transform into row-per seat
    nest(candidate = c(candidate,
                       display_name,
                       is_incumbent,
                       party,
                       street_address,
                       city_town,
                       statement)) %>%
    # Join to incumbency data
    left_join(running_incumbents, by=c("office", "district")) %>%
    # District-level data
    mutate(num_candidates = map_int(candidate, nrow),
           is_contested = (num_candidates > 1),
           is_open = is.na(incumbent),
           parties = map(candidate, party_counts)) %>%
    unnest(parties) %>%
    # This only handles D, R, and U for `party`
    replace_na(list(D=0, R=0, U=0)) %>%
    mutate(inc_party_count = case_when(
        incumbent_party == "D" ~ D,
        incumbent_party == "R" ~ R,
        incumbent_party == "U" ~ U,
        TRUE ~ NA_real_)) %>%
    mutate(incumbent_primary = (inc_party_count > 1),
           contested_no_primary = (is_contested & !incumbent_primary)) %>%
    left_join(district_summaries, by="district_id") %>%
    rows_patch(calculated_incumbents,
               by=c("office", "district"),
               unmatched="ignore") %>%
    filter(office %in% offices_to_cover)

## notable_districts <- districts %>%
##     filter(office %in% offices_to_cover,
##           (incumbent_primary | is_open | contested_no_primary)) %>%
##     mutate(candidate_names = map_chr(candidate,
##                                      ~paste(.x$display_name, collapse = ", ")))
##
## notable_districts %>% write_csv("notable_districts_2024.csv")

office_table_display <- function(df, office_group) {
    df %>%
        filter(office == office_group) %>%
        select(-office) %>%
        gt(groupname_col="row_group", process_md=TRUE) %>%
        tab_options(
            column_labels.hidden=TRUE,
            row_group.background.color = group_color) %>%
        sub_missing(missing_text="") %>%
        cols_align(columns="display_name", align="right") %>%
        opt_row_striping(row_striping=FALSE)
}
```

This overview covers U.S. Senate, U.S. Congress, State Senate, State
Represenative, and Governor's Council races. It divides the races into
incumbents with contested primaries, open seats where the current seat
holder is not pursuing their seat in 2024, and then races with
non-primary competition for the incumbent.


## Primary challenges

```{r}
incumbent_primaries <- districts %>%
    filter(office %in% offices_to_cover,
           incumbent_primary)

incumbent_primary_view <- incumbent_primaries %>%
    select(office, district, district_id, summary, incumbent_party, candidate) %>%
    unnest(candidate) %>%
    select(office,
           district,
           district_id,
           summary,
           display_name,
           statement,
           is_incumbent,
           incumbent_party,
           party) %>%
    filter(incumbent_party == party) %>%
    mutate(row_group = str_glue("**{office} - {district}**<br/>*{summary}*")) %>%
    arrange(district_id, desc(is_incumbent)) %>%
    select(office, row_group, display_name, statement)

num_primary_challenges <- nrow(incumbent_primaries)
```

### Governor's Council incumbent primaries

```{r}
office_table_display(incumbent_primary_view, "Governor's Council")
```

### State Senate incumbent primaries

```{r}
office_table_display(incumbent_primary_view, "State Senate")
```

### State Representative incumbent primaries

```{r}
office_table_display(incumbent_primary_view, "State Representative")
```

## Open Seats

```{r}
open_seats <- districts %>%
    filter(is_open)

## incumbents_not_running <- open_seats %>%
##     select(district,
##            office,
##            incumbent_display)

open_seats_view <- open_seats %>%
    select(office, district, summary, incumbent_display, candidate) %>%
    unnest(candidate) %>%
    select(office, district, summary, display_name, statement, party, incumbent_display) %>%
    group_by(office, district) %>%
    arrange(party) %>%
    select(-party) %>%
    ungroup() %>%
    mutate(row_group = str_glue("**{office} - {district}**<br/>*Incumbent {incumbent_display} {summary}*")) %>%
    select(-c(district, summary, incumbent_display))
```

There are `{r} nrow(open_seats)` seats where the sitting incumbent is
not running. There is a section in the header listing the incumbent
with other details. At least eight of the incumbents are running for a
different office or have been appointed to another governmental role.

### Open Governor's Council

```{r}
office_table_display(open_seats_view, "Governor's Council")
```

### Open State Senate

```{r}
office_table_display(open_seats_view, "State Senate")
```

### Open State Representative

```{r}
office_table_display(open_seats_view, "State Representative")
```

## Contested

```{r}
contested <- districts %>%
    filter(is_contested,
           inc_party_count == 1)

contested_view <- contested %>%
    select(office, district, candidate) %>%
    unnest(candidate) %>%
    select(office, district, display_name, statement, is_incumbent, party) %>%
    group_by(office, district) %>%
    arrange(desc(is_incumbent), party) %>%
    ungroup() %>%
    mutate(row_group = str_glue("**{office} - {district}**")) %>%
    select(-c(district, is_incumbent, party))

num_gc_contested <- nrow(filter(contested, office == "Governor's Council"))
num_state_sen_contested <- nrow(filter(contested, office == "State Senate"))
num_state_rep_contested <- nrow(filter(contested, office == "State Representative"))
```

There are `{r} nrow(contested)` seat where the incumbent does not have
a primary challenges, but faces one or more candidates who are
unenrolled or from a different party. Senator Elizabeth Warren will
face off against the winner of a three-way Republican primary. There
are also three Republicans vying to replace U.S. Representative
Stephen Lynch, while U.S. Representative Bill Keating will face
Republican Dan Sullivan who lost the 2022 GOP primary to Jesse Brown.

There are `{r} num_gc_contested` Governor's Council challenges,
`{r} num_state_sen_contested` State Senate challenges, and
`{r} num_state_rep_contested` State Rep. challenges in addition to
the `{r} num_primary_challenges` incumbents with primaries mentioned
earlier.

### U.S. Senate contested

```{r}
office_table_display(contested_view, "Senator in Congress")
```

### U.S. Congress contested

```{r}
office_table_display(contested_view, "Representative in Congress")
```

### Governor's Councile contested

```{r}
office_table_display(contested_view, "Governor's Council")
```

### State Senate contested

```{r}
office_table_display(contested_view, "State Senate")
```

### State Representative contested

```{r}
office_table_display(contested_view, "State Representative")
```

